|

### 2024-01-15 paper/paper.Rmd
finish distance desc

@@

Residuals, within regression analysis, represent the differences between fitted values and observed data points, capturing the unexplained elements in the regression model. The practice of plotting residuals, advocated by influential regression literature [@cook1982residuals; @draper1998applied; @belsley1980regression; @montgomery1982introduction], serves as a standard procedure in regression diagnostics. This visual examination is crucial for identifying deviations from the model assumptions like linearity, homoscedasticity, and normality.

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. Consider Figure \ref{fig:false-finding} as an example, the residuals display a triangular shape pointing to the left. While this might suggest heteroskedasticity, it's important to avoid over-interpreting the visual pattern. In this case, the fitted model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

A residual plot can exhibit various visual features, but it's crucial to recognize that some may arise from the characteristics of predictors and the inherent randomness of the error, rather than indicating a violation of model assumptions [@li2023plot]. The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the actual residual plot significantly differs visually from null plots, which are created using residuals generated from the null distribution. Typically, this is accomplished through the lineup protocol. In this approach, the real residual plot is embedded within a lineup alongside several null plots. If the real residual plot can be distinguished from the lineup, it provides evidence for rejecting the null hypothesis.

@@

There is a compelling need to develop an approach that alleviates people's workload by automating repetitive tasks and providing standardized results in a controlled environment. The large-scale evaluation of lineups is impractical without the use of technology and machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatterplot diagnostics. @wilkinson2005graph expanded on this work in 2005, introducing graph theoretic scagnostics, which encompassed nine computable measures applied to planar proximity graphs. These measures, including "Outlying," "Skinny," "Stringy," "Straight," "Monotonic," "Skewed," "Clumpy," and "Striated," aimed to characterize outliers, shape, density, trend, and coherence of the data.

While this approach has been inspiring, there is a recognition that it may not capture all the necessary visual features distinguishing actual residual plots from null plots. An alternative, more promising approach involves allowing computers to learn the visual feature extraction function. In essence, this means enabling computers to discern which visual features are significant for residual diagnostics and how to extract them. Modern computer vision models, a subset of artificial intelligence, present a compelling solution to this challenge.

Modern computer vision models often rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers leverage hierarchical patterns in data, downsizing and transforming images by summarizing information in a small space. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep].

Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models represents a relatively new field of study.

@@

@@

@@

@@

## Model evaluation

@@

## Case study 1: ...

## Case study 2: ...

## Case study 3: datasaurus

>>>>

<!-- Who is the audience? -->

<!-- someone cares about linear regression -> statistics journals -->

<!-- -> explain CNN design -->
<!-- ->  -->

@@

The practice of plotting residuals is commonly regarded as a standard procedure in linear regression diagnostics \citep[see][]{cook1982residuals, belsley1980regression}. This visual assessment plays a crucial role in identifying deviations from model assumptions, such as linearity, homoscedasticity, and normality. It also helps in understanding the goodness of fit and various characteristics of the model.

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. Consider Figure \ref{fig:false-finding} as an example, the residuals display a triangular shape pointing to the left. While this might suggest heteroskedasticity, it is important to avoid over-interpreting the visual pattern. In this case, the fitted model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the inherent randomness of the error, rather than indicating a violation of model assumptions [@li2023plot]. The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the actual residual plot visually differs significantly from null plots, which are created using residuals generated from the null distribution. Typically, this is accomplished through the lineup protocol. In this approach, the real residual plot is embedded within a lineup alongside several null plots. If the real residual plot can be distinguished from the lineup, it provides evidence for rejecting the null hypothesis.

@@

There is a substantial need to develop an approach that alleviates people's workload by automating repetitive tasks and providing standardized results in a controlled environment. The large-scale evaluation of lineups is impractical without the use of technology and machines.

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatterplot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed nine computable measures applied to planar proximity graphs. These measures, including "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, and coherence of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate actual residual plots from null plots. A more promising alternative entails enabling computers to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them. Modern computer vision models are well-suited for addressing this challenge.

Modern computer vision models often rely on deep neural networks with convolutional layers [@fukushima1982neocognitron]. These layers leverage hierarchical patterns in data, downsizing and transforming images by summarizing information in a small space. Numerous studies have demonstrated the efficacy of convolutional layers in addressing various vision tasks, including image recognition [@rawat2017deep]. Despite the widespread use of computer vision models in fields like computer-aided diagnosis [@lee2015image], pedestrian detection [@brunetti2018computer], and facial recognition [@emami2012facial], their application in reading data plots remains limited. While some studies have explored the use of computer vision models for tasks such as reading recurrence plots for time series regression [@ojeda2020multivariate], time series classification [@chu2019automatic; @hailesilassie2019financial; @hatami2018classification; @zhang2020encoding], anomaly detection [@chen2020convolutional], and pairwise causality analysis [@singh2017deep], the application of reading residual plots with computer vision models represents a relatively new field of study.

@@

## Distance from the good residual plots

In a visual test, the observer will be asked to choose one or more plots that stand out as most distinct from others in a given lineup. To develop a computer vision model for evaluating residual plots within the visual inference framework, it is important to precisely define a numerical measure of "difference" or "distance" between plots. This distance can take the form of a basic statistical operation on pixels, such as the sum of square differences. Alternatively, it could involve established image similarity metrics like the Structural Similarity Index Measure (SSIM) (ref here). The challenge lies in the fact that metrics tailored for image comparison may not be suitable for evaluating data plots, where only essential plot elements require assessment [@chowdhury2018measuring].

### Residual distribution

The distance metrics proposed in this paper takes into account the fact that we try to measure how different a residual plot is from a good residual plot, or in other words, how different a given fitted model is from a correctly specified model. For classical normal linear regression model, residuals are derived from the fitted values $\hat{\boldsymbol{y}}$ and observed values $\boldsymbol{y}$. However, if the data generating process is known and the model is correctly specified, by the Frisch-Waugh-Lowell theorem (ref here), residuals $\boldsymbol{e}$ can be written as a linear transformation of the error $\boldsymbol{\varepsilon}$:

\begin{equation}
\label{eq:residual-1}
\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon},
\end{equation}

\noindent where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

One of the assumptions of the classical normal linear regression model is the error $\boldsymbol{\varepsilon}$ follows a multivariate normal distribution with zero mean and constant variance, i.e., $N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$. Using equation \ref{eq:residual-1}, it can be known that residuals $\boldsymbol{e}$ also follow a certain probability distribution, which will be denoted as $Q$. This reference distribution $Q$ summarises what good residuals should follow given the predictors are known and fixed.

Probability distribution $Q$ is actually a degenerate multivariate distribution due to the fact that $rank(\boldsymbol{R}) = n - 1 < n$. This means the $n$th residual value will be known given the $n - 1$ residuals. To capture the exact characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$. For simplicity, we replaced $\boldsymbol{R}$ with a diagonal matrix $diag(\boldsymbol{R})$. The resulting distribution for $\boldsymbol{Q}$ is $N(\boldsymbol{0}_n, diag(\boldsymbol{R}\sigma^2))$.

If the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different to $Q$. For example, if the model formula is correct, but the error $\boldsymbol{\varepsilon}$ follows a multivariate lognormal distribution, $P$ will not be the same as $Q$ and we would expect a skewed empirical distribution for residuals. And if there is an omitted variable problem, then the residual operator obtained from the fitted model will not be the same as $\boldsymbol{R}$, and again, $P$ will be different to $Q$.

### Kullback-Leibler divergence of $P$ from $Q$

Define a proper distance between distributions is usually easier than define a proper distance between data plots. Given the actual residual distribution $Q$ and the reference residual distribution $P$, we used a distance metric based on Kullback-Leibler divergence (ref here) to quantify the difference between two distributions

\begin{align}
D &= log\left(1 + KL\right), \\
\label{eq:kl-1}
KL &= \int_{\mathbb{R}^{n}}log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{align}

\noindent where $p(\boldsymbol{e})$ is the probability density function for distribution $P$, and $q(\boldsymbol{e})$ is the probability density function for distribution $Q$.

This distance metric was first proposed in @li2023plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. For a linear regression model that has omitted higher-order predictors $\boldsymbol{Z}$ along with coefficients $\boldsymbol{\beta}_z$, and an error distribution $N(\boldsymbol{0}_n, \boldsymbol{V})$, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'))$. Note that the variance-covariance matrix is replaced with its diagonal analogy to ensure it is a full rank matrix.

Since both $P$ and $Q$ are adjusted to be multivariate normal distributions, equation \ref{eq:kl-1} can be further expanded to

\begin{align}
\label{eq:kl-2}
KL &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(diag(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'$. The assumed error variance $\sigma^2$ is set to be $tr(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

### Evaluation of Kullback-Leibler divergence for non-normal $P$

For non-normal error $\boldsymbol{\varepsilon}$, the actual residual distribution $P$ is unlikely to be a multivariate normal distribution. Thus, equation \ref{eq:kl-2} will not be applicable to linear regression model with non-normality issue.

To evaluate the Kullback-Leibler divergence of non-normal $P$ from $Q$, we need to solve equation \ref{eq:kl-1}. However, since $\boldsymbol{e}$ is a linear transformation of non-normal random variables, it is very common that the general form of $P$ is unknown, meaning that we can not easily compute $p(\boldsymbol{e})$ using a well-known probability density function.

Additionally, even if $p(\boldsymbol{e})$ can be calculated for any $\boldsymbol{e} \in \mathbb{R}^n$, it will be very difficult to do numerical integration over the $n$ dimensional space, because $n$ could be potentially very large.

In order to evaluate equation \ref{eq:kl-1} in a practically computable manner, the elements of $\boldsymbol{e}$ are assumed to be independent of each other. This assumption solves both of the issues mentioned above. First, we no longer need to integrate over $n$ random variables. The result is now the sum of the Kullback-Leibler divergence evaluated for each individual residual thanks for the independence assumption. Second, it is not required to know the joint probability density $p(\boldsymbol{e})$ any more. Instead, the evaluation of Kullback-Leibler divergence for a single residual relies on the knowledge of $p(e_i)$, where $e_i$ is the $i$th residual for $i = 1, ..., n$. This is much easier to estimate through simulation.

The algorithm to compute equation \ref{eq:kl-1} starts from simulating $m$ sets of $\boldsymbol{\varepsilon}$ according to the error distribution. The simulated errors are stored in a matrix $\boldsymbol{A}$ with $n$ rows and $m$ columns. So each column of $\boldsymbol{A}$ is a set of realization values of $\boldsymbol{\varepsilon}$. Then, we can get $m$ sets of $\boldsymbol{e}$ stored in the matrix $\boldsymbol{B}$ by applying the residual rotation operator $\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}$. Furthermore, kernel density estimation (ref here) is applied on each row of $B$ to estimate $p(e_i)$ for $i = 1, ..., n$.

Since the Kullback-Leibler divergence can be viewed as the expectation of the log-likelihood ratio between distribution $P$ and distribution $Q$ evaluated on distribution $P$, we can reuse the simulated residuals in matrix $\boldsymbol{B}$ to compute the expectation. Therefore, the Kullback-Leibler divergence can be formulated as

\begin{align}
\label{eq:kl-3}
KL &= \sum_{i = 1}^{n} KL_i, \\
KL_i &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{p(B_{ij})}{q(B_{ij})},
\end{align}

\noindent where $KL_i$ is the Kullback-Leibler divergence for an individual residual $e_i$, $B_{ij}$ is the $i$th row and $j$th column entry of the matrix $B$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\hat{\sigma^2} = var(flatten(B))$, and $flatten(.)$ unfolds a matrix into a vector.

### Approximation of the distance metric with a residual plot

In the previous sections, we have defined a distance metric for quantifying the difference between the actual residual distribution and an ideal reference distribution. This distance metric can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to fit a linear regression model in the first place.

What we proposed in this paper is a method to approximate this distance with a residual plot. This is done by training computer vision models with a large number of pairs of residual plots and distance. With the approximated distance, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. This is meaningful way to check if a residual plot is a good residual plot, and to know the strength of the visual signal indicating model violations.

The approximated distance is not expected to be the same as the original distance. This is largely because information embedded in a single residual plot is limited and it may not be able to summarise the characteristics of the residual distribution. For a given residual distribution $P$, we can generate many different residual plots. Some of them share similar visual patterns, but some of them could be visually very different from the rest, especially for models with small $n$. This suggests the approximated distance will vary depends on whether the observed residual plot is representative or not.

@@

The computer vision models we developed in this study were trained with synthetic data. This means the training data and the test data were simulated from a known data generating process under different parameter settings.

We have incorporated three types of residual departures of linear regression model in this study, including non-linearity, heteroskedasticity and non-normality.

<!-- The training data and test data was simulated using the following data   -->

<!-- The data was simulated using the following data generating process -->

\begin{align*}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= He_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= He_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2}, \\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}_n, \boldsymbol{I}_n),
\end{align*}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $He_j(.)$ is the $j$th-order probabilist's Hermite polynomials (ref), the $\sqrt{(.)}$ and $(.)^2$ operator are element-wise operators, $\odot$ is the Hadamard product, and $g(., k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{min}\boldsymbol{1}_n}{x_{max} - x_{min}} - k\boldsymbol{1}_n,~for~k > 0,$$
\noindent where $x_{min} = \underset{i \in \{ 1,...,n\}}{min} x_i$, $x_{max} = \underset{i \in \{ 1,...,n\}}{max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.

## Different configurations of the model formula

discuss different types of inputs and output choices, and why we choose the current design

@@

## Model evaluation methods

@@

## Best model performance

@@

## When the model works

- simple examples (non-linearity, heteroskedasticity, ...)
- datasaurus

## When the model does not work

- human detect but model does not
- cartoon residuals?

## Workflow: how one use this model? (small showcase)
